HO unification from object language to meta language

### 2024-04-02 paper/main.tex
visible TODO

\marginpar{fix}Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
is resilient to unification problems outside \llambda solved by
a third party. We believe this property is of practical interest since we
want the user to be able to add heuristics via hand written rules
to the ones obtained by our compilation scheme. A Typical example
is the following problem \ref{problem:q} that is outside \llambda:

@@

\noindent
If a memory cell is \elpiIn{none}, then the corresponding unification variable
is not set. \elpiIn{assign} sets an unset cell to the given value, while
\elpiIn{new} finds the first unused address and sets it to \elpiIn{none}.\marginpar{is new used?}

@@

\noindent
The only detail worth discussing is the fact that the procedure updates a
substitution, rather than just crafting one as presented in
section~\ref{sec:problem-statement}. The reason is that the algorithm folds
over a term, updating a substitution while it traverses it.\marginpar{explain better}

@@

% The solution we are proposing aim to overcome these unification issues by 1)
% compiling the terms $t$ and $u$ of the OL into an internal version $t'$ and $u'$
% in the ML; 2) unifying $t'$ and $u'$ at the meta level instantiating meta
% variables; 3) decompiling the meta variable into terms of the OL; 4) assigning
% the variables of the OL with the decompiled version of their corresponding meta
% variables. We claim that $t$ and $u$ unify if and only if $t'$ and $u'$ unify
% and that the substitution in the object language is the same as the one returned
% by the ML. \todo{same or $\supseteq$ or $\subseteq$}

@@

\subsection{Compilation}
\label{sec:compilation}
\marginpar{manca beta normal in entrata}

@@

% The following code
% %
% \begin{elpicode}
%   kind link type.
%   type link nat -> nat -> nat -> subst.
% \end{elpicode}
% %
% \noindent
% defines a link, which is a relation between to variables indexes, the first
% being the index of a OL variable and the second being the index of a ML
% variable. The third integer\todo{integer or nat?} is the number of term in the
% scope of the two variables, or equivalently, in a typed language, their arity.

@@

\marginpar{the compiler must do beta normal!!!!}
\newcommand{\reduceto}{\emph{may-contract-to}}
\begin{definition}[\reduceto]
  A term $s$ \reduceto{} a name $x$ if there exists a
  substitution $\rho$ such that $\rho s \Eo{} x$.
\end{definition}

@@

\noindent
Note that this condition does not require the term to be in \llambda.
\marginpar{Is this relevant}

@@

\newcommand{\progressetaleft}{\emph{progress-$\eta$-left}}
\begin{definition}[\progressetaleft]\label{def:progressetaleft}
A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
$X$ becomes rigid. There are two cases:
\begin{enumerate}
  \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
    we unify the $\eta$-expansion of the $X$ with $T$, that is we run
    $\lambda x.X \appsep x\Ue{} T$  (under
    the context $\Gamma$)
  \item if $X = \lambda x.t$ we run $X \Ue{} T$.
  \marginpar{do we have the name case in the code?}
\end{enumerate}
\end{definition}

\newcommand{\progressetaright}{\emph{progress-$\eta$-right}}
\begin{definition}[\progressetaright]\label{def:progressetaright}
A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
$\testmaybeeta~T$ does not hold (anymore) and by $\eta$-contracting
$T$ to $T'$ (if possible, else $T' = T$) and executing $X \Ue{} T'$ (under
the context $\Gamma$).
\marginpar{there was a note about T' having a rigid head, TO BE FIXED}
\end{definition}

@@

\paragraph{Example of \progressetaright}\marginpar{TODO}

@@

\marginpar{I don't get this TODO}
TODO: we can have $\lambda x. F_x$ in the substitution if we know that $F$ does
not reduce to $T x$ where x is not free in $T$.

@@

If we look back the example give at the beginning of this section, we can
deduplicate $\mapping{X}{E}{1}, \mapping{X}{C}{2}$ by removing the first mapping
and adding the auxiliary \linketa: \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
After deduplication the compiler output is as follows:
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{Y,F,0},
    {X,C,2}}}
  {{{\eta,x,E_{x},\lambda y.C_{x y}},
    {\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}
\marginpar{TODO: non so se serva spiegare l'esecuzione (in commento)}
% After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
% This assignment makes $l_2$ to progress since the \lhs is materialized and by
% unification, between $X$ and $\lambda x.Y_{x}$, $Y_x$ is instantiate to $\lambda
% y.x$. Once $Y_x$ is instantiated, $l_1$ can progress, and set $H_{xy}$ to $x$.
% After all these progresses, $l_1$ and $l_2$ are remove from \linkStore and the
% \elpiIn{progress} fixpoint terminates. Next, the second unification problem is
% run, and $Z$ is set to $f\appsep(\lambda x.x)$. This unification wakes up $l_3$
% and since $Z$ starts with the \elpiIn{app} node, the $\eta$-expanded version of
% $Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
% As last step, the last link is progressed and the final \Ho substitution is
% $\{X _{} \mapsto \lambda x.\lambda y.x, Y _{x} \mapsto \lambda y.x, 
% G _{yx} \mapsto y, Z _{} \mapsto f~\lambda x.x, 
% H _{x} \mapsto \lambda y.y\}$.
% The decompilation phase is only charged, in this example to solve the mappings,
% since no suspended links remain. The only mapping in the list is
% \mapping{F}{H}{2}, which will assign the $F$ variable in \Fo{} to $\lambda xy.y$

@@

Until now, we have only dealt we unification of terms in \llambda. However, we
want the unification relation to be more robust so that it can work with terms
in \maybebeta. Unification in \maybebeta is in general a non-decidable
procedure, e.g. $X \appsep a \Uo a$ is a unification problem in \maybebeta,
since we have $X a$ has the shape $X t_1\dots t_n$ where $X$ is a unification
variable and $t_1\dots t_n$ is not a list of distinct names (in our example, $a$
is a constant, hence not a name). We also point out that this unification
problem admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
and $\rho_2 = \{X \mapsto \lambda \_.a\}$.

@@

In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
$\foUnifPb_2$, can be solved in \llambda. On the other hand, we see that, 
\Ue can't solve the compiled problems \hoUnifPb. In
fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
\mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
non-unifable problem $(\lambda x. B) \appsep a \Ue a$.

In order to encompass this unification, term compilation should capture the
terms $t$ in \maybebeta and replace them with fresh varaibles $v$. As per
\maybebeta, the variables $v$ and the terms $t$ are linked through a \linkbeta.

>>>>

\todo[inline]{fix}Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
is resilient to unification problems outside \llambda solved by
a third party. We believe this property is of practical interest since we
want the user to be able to add heuristics via hand written rules
to the ones obtained by our compilation scheme. A Typical example
is the following problem \ref{problem:q} that is outside \llambda:

@@

\noindent
If a memory cell is \elpiIn{none}, then the corresponding unification variable
is not set. \elpiIn{assign} sets an unset cell to the given value, while
\elpiIn{new} finds the first unused address and sets it to \elpiIn{none}.\todo[inline]{is new used?}

@@

\noindent
The only detail worth discussing is the fact that the procedure updates a
substitution, rather than just crafting one as presented in
section~\ref{sec:problem-statement}. The reason is that the algorithm folds
over a term, updating a substitution while it traverses it.\todo[inline]{explain better}

@@

% The solution we are proposing aim to overcome these unification issues by 1)
% compiling the terms $t$ and $u$ of the OL into an internal version $t'$ and $u'$
% in the ML; 2) unifying $t'$ and $u'$ at the meta level instantiating meta
% variables; 3) decompiling the meta variable into terms of the OL; 4) assigning
% the variables of the OL with the decompiled version of their corresponding meta
% variables. We claim that $t$ and $u$ unify if and only if $t'$ and $u'$ unify
% and that the substitution in the object language is the same as the one returned
% by the ML. \todo[inline]{same or $\supseteq$ or $\subseteq$}

@@

\subsection{Compilation}
\label{sec:compilation}
\todo[inline]{manca beta normal in entrata}

@@

% The following code
% %
% \begin{elpicode}
%   kind link type.
%   type link nat -> nat -> nat -> subst.
% \end{elpicode}
% %
% \noindent
% defines a link, which is a relation between to variables indexes, the first
% being the index of a OL variable and the second being the index of a ML
% variable. The third integer\todo[inline]{integer or nat?} is the number of term in the
% scope of the two variables, or equivalently, in a typed language, their arity.

@@

\todo[inline]{the compiler must do beta normal!!!!}
\newcommand{\reduceto}{\emph{may-contract-to}}
\begin{definition}[\reduceto]
  A term $s$ \reduceto{} a name $x$ if there exists a
  substitution $\rho$ such that $\rho s \Eo{} x$.
\end{definition}

@@

\noindent
Note that this condition does not require the term to be in \llambda.
\todo[inline]{Is this relevant}

@@

\newcommand{\progressetaleft}{\emph{progress-$\eta$-left}}
\begin{definition}[\progressetaleft]\label{def:progressetaleft}
A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
$X$ becomes rigid. There are two cases:
\begin{enumerate}
  \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
    we unify the $\eta$-expansion of the $X$ with $T$, that is we run
    $\lambda x.X \appsep x\Ue{} T$  (under
    the context $\Gamma$)
  \item if $X = \lambda x.t$ we run $X \Ue{} T$.
  \todo[inline]{do we have the name case in the code?}
\end{enumerate}
\end{definition}

\newcommand{\progressetaright}{\emph{progress-$\eta$-right}}
\begin{definition}[\progressetaright]\label{def:progressetaright}
A link \linketaM{\Gamma}{X}{T} is removed from \linkStore when
$\testmaybeeta~T$ does not hold (anymore) and by $\eta$-contracting
$T$ to $T'$ (if possible, else $T' = T$) and executing $X \Ue{} T'$ (under
the context $\Gamma$).
\todo[inline]{there was a note about T' having a rigid head, TO BE FIXED}
\end{definition}

@@

\paragraph{Example of \progressetaright}\todo[inline]{TODO}

@@

\todo[inline]{I don't get this TODO}
TODO: we can have $\lambda x. F_x$ in the substitution if we know that $F$ does
not reduce to $T x$ where x is not free in $T$.

@@

If we look back the example give at the beginning of this section, we can
deduplicate $\mapping{X}{E}{1}, \mapping{X}{C}{2}$ by removing the first mapping
and adding the auxiliary \linketa: \linketaM{x}{E_{x}}{\lambda y.C_{x y}}.
After deduplication the compiler output is as follows:
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{Y,F,0},
    {X,C,2}}}
  {{{\eta,x,E_{x},\lambda y.C_{x y}},
    {\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}
\todo[inline]{TODO: non so se serva spiegare l'esecuzione (in commento)}
% After unification of the two terms, $X$ is assigned to $\lambda x.\lambda y.x$.
% This assignment makes $l_2$ to progress since the \lhs is materialized and by
% unification, between $X$ and $\lambda x.Y_{x}$, $Y_x$ is instantiate to $\lambda
% y.x$. Once $Y_x$ is instantiated, $l_1$ can progress, and set $H_{xy}$ to $x$.
% After all these progresses, $l_1$ and $l_2$ are remove from \linkStore and the
% \elpiIn{progress} fixpoint terminates. Next, the second unification problem is
% run, and $Z$ is set to $f\appsep(\lambda x.x)$. This unification wakes up $l_3$
% and since $Z$ starts with the \elpiIn{app} node, the $\eta$-expanded version of
% $Z$ is unified with $\lambda x.f \appsep G_{x}~x$ and $G_x$ is set to $x$.
% As last step, the last link is progressed and the final \Ho substitution is
% $\{X _{} \mapsto \lambda x.\lambda y.x, Y _{x} \mapsto \lambda y.x, 
% G _{yx} \mapsto y, Z _{} \mapsto f~\lambda x.x, 
% H _{x} \mapsto \lambda y.y\}$.
% The decompilation phase is only charged, in this example to solve the mappings,
% since no suspended links remain. The only mapping in the list is
% \mapping{F}{H}{2}, which will assign the $F$ variable in \Fo{} to $\lambda xy.y$

@@

\todo[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions. Third order, functions that take functions, is undecidable, proved by Dowek. IMO here the problem is that there is no unique solution, (semi)decidability is a secondary point. This paragraph needs fixing.}
Until now, we have only dealt we unification of terms in \llambda. However, we
want the unification relation to be more robust so that it can work with terms
in \maybebeta. Unification in \maybebeta is in general a non-decidable
procedure, e.g. $X \appsep a \Uo a$ is a unification problem in \maybebeta,
since we have $X a$ has the shape $X t_1\dots t_n$ where $X$ is a unification
variable and $t_1\dots t_n$ is not a list of distinct names (in our example, $a$
is a constant, hence not a name). We also point out that this unification
problem admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
and $\rho_2 = \{X \mapsto \lambda \_.a\}$.

@@

In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
$\foUnifPb_2$, can be solved in \llambda.
\todo[inline]{it is even a ground term, there is no unification left to perform actually}
On the other hand, we see that, 
\Ue can't solve the compiled problems \hoUnifPb. In
fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
\mapsto \lambda x. B\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
non-unifable problem $(\lambda x. B) \appsep a \Ue a$.

In order to
encompass\todo[inline]{not sure has the right meaning}
this unification, term compilation should capture the
terms $t$ in \maybebeta and replace them with fresh varaibles $v$.
\todo[inline]{we use capitals for unif variables}
As per
\maybebeta, the variables $v$ and the terms $t$ are linked through a \linkbeta.
